---
title: "Keystone Exercise"
author: "Caterina Palma"
date: "'r format(Sys.time(), '%d %B, %Y')'"
#output: 
html_document: 
  code_folding: show 
  collapse: no
  df_print: paged 
  fig_caption: yes
  fig_height: 6 
  fig_width: 10 
  highlight: textmate 
  theme: spacelab 
  toc: yes
  toc_float: yes 
editor_topions: 
chunk_output_type: inline 
---

#Introduction 

Here we are going to wrangle messy data on the Ross River estuary fish surveys and produce clean tables and figure. 
We have 4 seperate files to wrangle in and use. 

```{r setup, include=FALSE}
#Set options, housekeeping
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

#Housekeeping 
```{r housekeeping }
#| message: true
#| warning: false
rm(list = ls())
```

#Packages 
```{r}
library(tidyverse) #includes dplyr, ggplot2, lubridate etc. 
library(readxl) #for reading in excel files 
```

#Main dataset
```{r}
#getwd() #this helps me see where my file path is. 
#The catch data is split across 4 table in one excel file, so we load each tab/sheet individually, hence sheet = 1, sheet = 2 etc. 

catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 2)
catch3 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 3)
catch4 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 4)
```

#Alternative method for loading files 
```{r}
#the file is anchored where the .qmd file is 
file_path <- "data/workshop2/estuary_catch_log.xlsx"
catch_data <- excel_sheets(file_path) |> 
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

#Inspect individual tabs 
```{r}
catch1
catch2
catch3
catch4
```

#Binding Rows Together 
```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
#clean up the environment (i.e. removes the catch data from our environment window)
rm(catch1, catch2, catch3, catch4)
```

#Examine data
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#Looking at Species and Site names: 
Can't tell from these outputs how many levels of site and species we have. 
Two options: let's check those explicitly by using distinct() or can convert to factors. 
Let's look at the distinct values first, as we might need to clean before converting to factor vars. 
```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)

#Need to clean this dataset. Clean the names of species and site, we have 4 of each. 
```
Ok, so we have four species in there as well as four sites, which look to be located at different points on the Ross River. So, we need to: 
- Fix site names and place in reasonable order. 
- Can decide this now or we can load the metadata and ensure consistency. 
- Fix species names - to do this we'll need to make consistent and then use the taxonomic dictionary to translate to scientific names. 

#Load metadata 
```{r}
metadata <- read.csv("data/workshop2/estuary_metadata.csv")
```
Inspect the data... 
```{r}
head(metadata)
glimpse(metadata)
```
#Loading Taxonomic Dictionary .csv file: 
```{r}
species_dict <- read.csv("data/workshop2/species_dictionary.csv")
head(species_dict)
```
#Loading Sonde Data .csv file 
```{r}
sonde_data <- read.csv("data/workshop2/estuary_sonde_data.csv")
head(sonde_data)
```

Site names are all over the place, names aren't all similar. 
#NOTES: in the catch_data dataframe we are changing the site names so they are all the same genre, changing chcracter to be factor variable AND changing species name to all be the same genre. 
```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ","_")) |>
  #Convert all chr to fct
  mutate(across(where(is.character),as_factor)) |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_"))
```

Adding another column called scientific_name by using the key variable (the similar variable of species). 
```{r}
catch_data <- catch_data |> 
  left_join(species_dict, by = join_by(species == common_name))
#If I wanted to, I can delete the common name from the dataframe by using the below: 
```

```{r}
metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |>
  #Convert all chr to fct 
  mutate(across(where(is.character), as_factor))
```

#Recheck 
```{r}
catch_data |>
  str()
glimpse(metadata)
```
#NOTES: making a new dataframe called sensor_clean, using the sonde_data dataframe so I can change/clean the datetime, turbidity and salinity variables. 
```{r}
sensor_clean <- sonde_data |>
  mutate(
    #clean the messy character string (Dat/Month/Year Hour:Minute)
    datetime = dmy_hm(timestamp),
    #convert the -999.0 to NA 
    turbidity = na_if(turbidity, -999),
    #convert the negative values of salinity to NA 
    salinity = ifelse(salinity < 0, NA, salinity)
  )
```

#Making a new dataframe called sensor_clean_daily, which is going to have the mean of temperature, salinity and turbidity every 15 minutes into a day. 
```{r}
sensor_clean_daily <- sensor_clean |>
    mutate(
    date = floor_date(datetime, "day"), #rounds the dates down to just the date, without the minutes 1 (i.e. 1/5/26)
    date = as_date(date) #ensure it matches the date format of the catch_data log
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop"
  )
```


#Implicit zero observations i.e. did not include zero into the data, meaning didn't catch any fish
4x species 
4x sites 
30 surveys 
#We should have 480 observations, not 427. 
```{r}
4*4*30
```

#TO DO: 
1. Join the catch_data with metadata and sensor_clean data frames. 
2. Deal with implicit zero observations. 

#1. Done below: 
```{r}
catch_data <- catch_data |> 
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_clean_daily, by = join_by(site,date))
```

#2. Done below: We not have 480 rows. 
```{r}
catch_data <- catch_data |>
  complete(
    nesting(site, date, lat, lon, zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name) |>
  mutate(count = coalesce(count, 0))
```

#Median Salinity & Standard Error of Mean Salinity: 
This shows that the salinity is different in each zone i.e. less salty in upstream (near freshwater) than marine. 

```{r}
salinity_summary <- catch_data |>
  group_by(zone) |>
  summarise(
    sem_salinity = sd(mean_salinity)/sqrt(n()), #sem_salinity is standard error of mean salinity
    #sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups ="drop"
  )
```

#Or with plotrix: 
Here we have grouped by zone because again we are just looking at salinity, and the main variable we need to look at would be zone, because that is where the change will be. 
```{r}
library(plotrix)
catch_data |> 
  group_by(zone) |>
  summarise(
    sem_salinity = std.error(count),
    #sd_salinity = sd(mean_salinity, na.rm = TRUE)
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```
#Making Salinity Figure
```{r}
salinity_summary |>
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sem_salinity, ymax = median_salinity + sem_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation error bars",
    x = "Zone",
    y = "Median Salintiy (ppt)")
```
#Calculating Fish Stats 
```{r}
fish_stats <- catch_data |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count)/sqrt(n()),
    .groups = "drop"
  )
```

#Creating Fish Stats Figure (ggplot)
```{r}
fish_stats |>
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_point(position = "dodge", size = 2, pch = 23) +
  #geom_line(group = 1) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch),
                position = position_dodge(0.9), width = 0.2) +
  #add in points from master data 
  geom_point(data = catch_data, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  facet_wrap(~scientific_name, scales = "fixed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey0"),
    #legend.position = "bottom" 
    legend.position = "none"
  ) + 
  labs(
    #title = "Predatory Fish Abundance across Esturine Salinity Gradients", 
    #subtitle = "Ross River Estuary System, May 2026", 
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```