---
title: "Keystone Exercise"
author: "Caterina Palma"
date: "'r format(Sys.time(), '%d %B, %Y')'"
#output: 
html_document: 
  code_folding: show 
  collapse: no
  df_print: paged 
  fig_caption: yes
  fig_height: 6 
  fig_width: 10 
  highlight: textmate 
  theme: spacelab 
  toc: yes
  toc_float: yes 
editor_topions: 
chunk_output_type: inline 
---

#Introduction 

Here we are going to wrangle messy data on the Ross River estuary fish surveys and produce clean tables and figure. 
We have 4 seperate files to wrangle in and use. 

```{r setup, include=FALSE}
#Set options, housekeeping
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

#Housekeeping 
```{r housekeeping }
#| message: true
#| warning: false
rm(list = ls())
```

#Packages 
```{r}
library(tidyverse) #includes dplyr, ggplot2, lubridate etc. 
library(readxl) #for reading in excel files 
```

#Main dataset
```{r}
#getwd() #this helps me see where my file path is. 
#The catch data is split across 4 table in one excel file, so we load each tab/sheet individually, hence sheet = 1, sheet = 2 etc. 

catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 2)
catch3 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 3)
catch4 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 4)
```

#Alternative method for loading files 
```{r}
#the file is anchored where the .qmd file is 
file_path <- "data/workshop2/estuary_catch_log.xlsx"
catch_data <- excel_sheets(file_path) |> 
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

#Inspect individual tabs 
```{r}
catch1
catch2
catch3
catch4
```

#Binding Rows Together 
```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
#clean up the environment (i.e. removes the catch data from our environment window)
rm(catch1, catch2, catch3, catch4)
```

#Examine data
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#Looking at Species and Site names: 
Can't tell from these outputs how many levels of site and species we have. 
Two Options: 
1. Use distinct(site) to check how many different levels of site there are and do the same with distinct(species). 
2. Convert to factors. 
Let's look at the distinct values first, as we might need to clean before converting to factor variables. 
```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)
```
Ok, so we have four species and four sites, which look to be located at different points on the Ross River (zones). So, we need to: 
- Fix site names. 
- Fix species names - to do this we'll need to make consistent and then can use the taxonomic dictionary to translate to scientific names. 

#Load metadata 
```{r}
metadata <- read.csv("data/workshop2/estuary_metadata.csv")
```
Inspect the data... 
```{r}
head(metadata)
glimpse(metadata)
```
#Loading Taxonomic Dictionary .csv file: 
```{r}
species_dict <- read.csv("data/workshop2/species_dictionary.csv")
head(species_dict)
```
#Loading Sonde Data .csv file - which is sensor data. 
```{r}
sonde_data <- read.csv("data/workshop2/estuary_sonde_data.csv")
head(sonde_data)
```

#Changing site AND species names for catch_data data frame. 

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)), #all site names into lowercase 
    site = str_replace_all(site, " ","_")) |> #replace spaced with _
  #Convert all chr to fct
  mutate(across(where(is.character),as_factor)) |>
  mutate(
    species = str_to_lower(str_trim(species)), #converting all species names to lowercase 
    species = str_replace_all(species, " ", "_")) #adding _ instead of spaces 
```
Now all our site names and species names will be the same genre, lowercase and _. 

#Changing common name to scientific name using species_dict data frame 
```{r}
catch_data <- catch_data |> 
  left_join(species_dict, by = join_by(species == common_name))
```

#Changing site names for metadata data frame 
```{r}
metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |>
  #Convert all chr to fct 
  mutate(across(where(is.character), as_factor))
```
Now all site names are lowercase and have _ instead of a space. 

#Recheck 
```{r}
catch_data |>
  str()
glimpse(metadata)
```

#New data frame for clean sonde_data or sensor data
We are making a new data frame called sensor_clean, uing the sonde_data data frame, so we can clean/change the datetime, turbidity and salinity variables. 
```{r}
sensor_clean <- sonde_data |>
  mutate(
    #clean the messy character string (Dat/Month/Year Hour:Minute)
    datetime = dmy_hm(timestamp),
    #convert the -999.0 to NA 
    turbidity = na_if(turbidity, -999),
    #convert the negative values of salinity to NA 
    salinity = ifelse(salinity < 0, NA, salinity)
  )
```

#Daily data frame for sensor data. 
We are making a new data frame using sensor_clean, which is instead of having data from every 15 minutes of the day, it will just be the mean temperature, salinity and turbidity of day. Starting from may 1st, to may 30th 2026. 
```{r}
sensor_clean_daily <- sensor_clean |>
    mutate(
    date = floor_date(datetime, "day"), #rounds the dates down to just the date, without the minutes 1 (i.e. 1/5/26)
    date = as_date(date) #ensure it matches the date format of the catch_data log
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop"
  )
```

#Implicit zero observations 
The catch_data frame did not have the zero observations i.e. didn't catch any of the fish, so we will need to add it. 
- 4x species 
- 4x sites 
- 30 surveys 
We should have 480 observations, not 427. 
```{r}
4*4*30
```

#TO DO: 
1. Join the catch_data with metadata and sensor_clean data frames. 
2. Deal with implicit zero observations. 

#1. Done below: 
```{r}
catch_data <- catch_data |> 
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_clean_daily, by = join_by(site,date))
```

#2. Done below: We not have 480 rows. 
```{r}
catch_data <- catch_data |>
  complete(
    nesting(site, date, lat, lon, zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name) |>
  mutate(count = coalesce(count, 0))
```

#Median Salinity & Standard Error of Mean Salinity: 
This shows that the salinity is different in each zone i.e. less salty in upstream (near freshwater) than marine. 

```{r}
salinity_summary <- catch_data |>
  group_by(zone) |>
  summarise(
    sem_salinity = sd(mean_salinity)/sqrt(n()), #sem_salinity is standard error of mean salinity
    #sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups ="drop"
  )
```

#Or with plotrix: 
Here we have grouped by zone because again we are just looking at salinity, and the main variable we need to look at would be zone, because that is where the change will be. 
```{r}
library(plotrix)
catch_data |> 
  group_by(zone) |>
  summarise(
    sem_salinity = std.error(count),
    #sd_salinity = sd(mean_salinity, na.rm = TRUE)
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```
#Making Salinity Figure
```{r}
salinity_summary |>
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sem_salinity, ymax = median_salinity + sem_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation error bars",
    x = "Zone",
    y = "Median Salintiy (ppt)")
```
#Calculating Fish Stats 
```{r}
fish_stats <- catch_data |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count)/sqrt(n()),
    .groups = "drop"
  )
```

#Creating Fish Stats Figure (ggplot)
```{r}
fish_stats |>
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_point(position = "dodge", size = 2, pch = 23) +
  #geom_line(group = 1) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch),
                position = position_dodge(0.9), width = 0.2) +
  #add in points from master data 
  geom_point(data = catch_data, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  facet_wrap(~scientific_name, scales = "fixed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey0"),
    #legend.position = "bottom" 
    legend.position = "none"
  ) + 
  labs(
    #title = "Predatory Fish Abundance across Esturine Salinity Gradients", 
    #subtitle = "Ross River Estuary System, May 2026", 
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```